---
title: 'HW 5: Correlations and linear models. Tests for categorial variables'
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(vcd) # for mosaic plots
```

## 1. Vowel reduction in Russian
Pavel Duryagin ran an experiment on perception of vowel reduction in Russian language. The dataset `shva` includes the following variables:  
_time1_ - reaction time 1  
_duration_ - duration of the vowel in the stimuly (in milliseconds, ms)  
_time2_ - reaction time 2  
_f1_, _f2_, _f3_ - the 1st, 2nd and 3rd formant of the vowel measured in Hz (for a short introduction into formants, see [here](https://home.cc.umanitoba.ca/~krussll/phonetics/acoustic/formants.html))  
_vowel_ - vowel classified according the 3-fold classification (_A_ - _a_ under stress, _a_ - _a/o_ as in the first syllable before the stressed one, _y_ (stands for shva) - _a/o_ as in the second etc. syllable before the stressed one or after the stressed syllable, cf. _g_[_y_]_g_[_a_]_t_[_A_]_l_[_y_] _gogotala_ `guffawed').  
In this part, we will ask you to analyse correlation between f1, f2, and duration.
The dataset is available [https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/duryagin_ReductionRussian.txt](here).

### 1.0 
Read the data from file to the variable `shva`.
```{r 1.0}
shva <- read.csv('https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/duryagin_ReductionRussian.txt', sep="\t")
```

### 1.1 
Scatterplot `f1` and `f2` using `ggplot()`. 
Design it to look like the [following](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin1.png).
```{r 1.1}
ggplot(shva, aes(x=f1, y=f2, color=vowel)) +
  geom_point() + 
  labs(title='f2 and f1 of the reduced and stressed vowels')

```

### 1.2 
Plot the boxplots of `f1` and `f2` for each vowel using `ggplot()`.
Design it to look like [this](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin2.png) and [this](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin3.png).
```{r 1.2}
# f1 boxplot

ggplot(shva, aes(x=vowel, y=f1, fill = vowel)) + 
  geom_boxplot() +
  coord_flip()

# f2 boxplot

ggplot(shva, aes(x=vowel, y=f2, fill = vowel)) + 
  geom_boxplot() +
  coord_flip()
```

### 1.3 
Calculate Pearson's correlation of `f1` and `f2` (all data)
```{r 1.3}
cor.test(shva$f1, shva$f2)
```

### 
1.4 Calculate Pearson's correlation of `f1` and `f2` for each vowel
```{r 1.4}
with(subset(shva, vowel == 'A'), cor.test(f1, f2))

with(subset(shva, vowel == 'a'), cor.test(f1, f2))

with(subset(shva, vowel == 'y'), cor.test(f1, f2))

```

## 2 Linear regressions

### 2.1.1 
Use the linear regression model to predict `f2` by `f1`.
```{r 2.1.1}
model1 <- lm(data = shva, f2 ~ f1)
summary(model1)

```

### 2.1.2 
Write down the equation for f2 using coefficients from the model (e.g. $y =  b + kx$)
```{r 2.1.2}
y = 1639.70215 - 0.42875*x
```

### 2.1.3 
Provide the adjusted R$^2$
```{r 2.1.3}
0.3319 
```

### 2.1.4 
Add the regression line in the scatterplot 1.1.
```{r 2.1.4}
ggplot(shva, aes(x=f1, y=f2)) +
  geom_point(aes(color=vowel)) +
  geom_smooth(method=lm) +
  labs(title='f2 and f1 of the reduced and stressed vowels')
```

### 2.1.5 
Make a scatter plot for `f1` and `f2` grouped by vowels. 
Use `ggplot()` and `facet_wrap()`.

```{r 2.1.5}
ggplot(shva, aes(x=f1, y=f2, color = vowel)) +
  geom_point() +
  facet_wrap(~vowel)
```

### 2.2.1 
Use the linear regression model to predict `f2` by `f1` and `vowel`.
```{r 2.2.1}
model2 <- lm(data = shva, f2 ~ f1 + vowel)
summary(model2)
```

### 2.2.2 
What is the intercept of the model?
```{r 2.2.2}
1365.91639
```

### 2.2.3 
Provide the adjusted R$^2$
```{r 2.2.3}
0.4801
```

### 2.2.4 
Write down your general conclusions about the relationship between `f1`, `f2`, and `vowels`.

```
If in the model we take into account not only the value of f1, but also the values of the variable 'vowel', then the p-value for f1 is less than the significance level, so we can conclude that the values of f1 have a weak effect on what value of f2 we get. The value of f2 mainly depends not on f1, but on the value of the categorical variable 'vowel'.

If 'vowel' takes the value a, then f2 on average takes the value 1365.91639
If 'vowel' takes the value A, then f2 on average takes the value 1365.91639 - 117.39221 = 1248.52
If 'vowel' takes the value y, then f2 on average takes the value 1365.91639 + 42.06747 = 1407.98
```

## 3. Dutch causative constructions

When the Dutch use two near-synonymous periphrastic causative verbs, *doen* and *laten*?
```
       De politie deed/liet de auto stoppen.
  lit. the police did/let the car stop 
       'The police stopped the car'
```

This is a data set on two rival constructions with *doen* and *laten* sampled from the newspaper corpora. The data frame includes 500 observations on the following 7 variables:

* `Aux` -- verb: a factor with levels `doen` and `laten`  
* `CrSem` -- the semantic class of the Causer: a factor with levels `Anim` (animate) and `Inanim` (inanimate)  
* `CeSem` -- the semantic class of the Causee: a factor with levels `Anim` (animate) and `Inanim` (inanimate)  
* `CdEvSem` -- the semantic domain of the caused event expressed by the Effected Predicate: a factor with levels `Ment` (mental) and `NonMent` (e.g. physical or social)  
* `CeSynt` -- the syntactic status of the Causee: a factor with levels `Clause`, `Impl` (implicit, not expressed), `NP` (noun phrase), `PP` (prepositional phrase)    
* `EPTrans` -- transitivity or intransitivity of the effected predicate, a factor with two levels `Tr` and `Intr`  
* `Country` -- a factor with levels `BE` (Belgium) and `NL` (Netherlands)  
* `Domain` -- a factor with four levels for newspaper domains.    

Data from Natalya Levshina's `RLing` package available (here)[https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/dutch_causatives.csv]
Read more on the constructions in [Levhina, Geerarts, Speelman 2014](https://www.academia.edu/7287585/Dutch_causative_constructions_Quantification_of_meaning_and_meaning_of_quantification_with_Dirk_Geeraerts_and_Dirk_Speelman_).

### 3.0 
Read the data from file to the variable `d_caus`.
```{r 3.0}
d_caus <- read.csv("https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/dutch_causatives.csv")
summary(d_caus)
```

### 3.1 
We are going to test whether the association between `Aux` and other categorical variables (`Aux` ~ `CrSem`, `Aux` ~ `CeSem`, etc) is statistically significant. The assiciation with which variable should be analysed using Fisher's Exact Test and not using Pearson's Chi-squared Test? Is this association statistically significant?
```{r 3.1}
# We should use Fisher's Exact Test to analyse association between Aux and CeSynt because there is value<5 among expected frequencies
fisher.test(d_caus$Aux, d_caus$CeSynt)
```

```
The association is statistically significant.
```

### 3.2. 
Test the hypothesis that `Aux` and `EPTrans` are not independent with the help of Pearson's Chi-squared Test. 
```{r 3.2}
res <- chisq.test(d_caus$Aux, d_caus$EPTrans)
res
```

```
p-value < 0.001 so we can claim that `Aux` and `EPTrans` are dependent
```

### 3.3 
Provide expected frequencies for Pearson's Chi-squared Test of `Aux` and `EPTrans` variables.
```{r 3.3}
res$expected
```

### 3.4. 
Calculate the odds ratio for observed frequencies of `Aux` and `EPTrans`
For 2×2 contigency table
$$\begin{matrix}
a & b \\
c & d
\end{matrix}
$$
one can find *odds ratio* as $(a/c)/(b/d)$.

```{r 3.4}
res$observed
odds_ratio <- (57/182)/(28/233)
odds_ratio

```

### 3.4.1 
Find odds ratio for expected frequencies of `Aux` and `EPTrans`

```{r 3.4.1}
exp_odds_ratio <- (40.63/198.37)/(44.37/216.63)
exp_odds_ratio
```
What can you say about odds ratio of expected frequencies for arbitrary data?

```
It must always be equal to one as it represents lack of association between variables.
```

### 3.5 
Calculate effect size for this test using Cramer's V (phi).
```{r 3.5}
tab <- table(d_caus$Aux, d_caus$EPTrans)
cramersV(tab)
```

### 3.6. 
Report the results of independence test using the following template:
```
We have found a significant association between variables `Aux` and `EPTrans` (p < 0.001).  The odds of ratios were 2.6 times higher in group Intr than in group Tr. Effect size is small (Cramer's V = 0.1691587).
```

### 3.7 
Visualize the distribution using mosaic plot.
Use `mosaic()` function from `vcd` library.
```{r 3.7}
mosaic(data=d_caus, EPTrans ~ Aux)
```

Below is an example of how to use mosaic() with three variables.
```{r 3.7.1}
#mosaic(~ Aux + CrSem + Country, data=d_caus, shade=TRUE, legend=TRUE)
```

### 3.8 
Why is it not recommended to run multiple Chisq tests of independence on different variables within your dataset whithout adjusting for the multiplicity? (i.e. just testing all the pairs of variables one by one)  
```
With multiple tests, the probability of mistakenly rejecting the null hypothesis increases significantly. The probability of rejecting the hypothesis with n tests is 1- (1-a) ^ n. That's why one should run multiple tests with adjusting for the multiplicity.
```

### 3.9 
Provide a short text (300 words) describing the hypothesis of this study and the results of your analysis.
```
The purpose of this study was to find out whether there is a statistically significant association between the use of a verb ('doen' or 'laten') and some other parameters (the semantic class of the Causer and Causee, the syntactic status of the Causee etc). In this case, the null hypothesis is that such an association does not exist and our variables are independent. H1 is that there is an association and variables are dependent. To identify the association between the two categorical variables, we use the chi-square test. As an example, we conduct a test for the variables `Aux` and` EPTrans`, which contains information about transitivity or intransitivity of the effected predicate. The chi-square test for these two variables gives a p-value of 0.0001553, which is less than 0.001, which allows us to reject the null hypothesis and claim that `Aux` and` EPTrans` are dependent. At the same time, the value of Cramer's V is 0.1691587, which for two degrees of freedom means small effect size (0.169 <0.21), so, according to this value, association between this two variables is not very strong.In order to complete the study and make a final conclusion about the relationship between the use of the verb 'doen' or 'laten' and other parameters, we can conduct tests for all other variables from our data as well.However, making the final conclusion, we must take into account that in the case of multiple tests, the probability of erroneously rejecting the null hypothesis increases, so it will be necessary to apply adjusting for the multiplicity.

```
